---
title: "Reproducible Research: Peer Assessment 1"
author: LexoL
date: January 10, 2016
output: 
  html_document:
    keep_md: true
---

The report falls into 8 sections that corresponds the first eight steps of the course project assignment. The ninth step neen not be presented in this report. 

## 1. Code for reading in the dataset and/or processing the data

First, we check if all the libraies and file are avilable. 

```{r, echo = TRUE}
if(!require(dplyr, quietly = TRUE, warn.conflicts = FALSE)){
        stop('I need dplyr library. Please run install.packages("dplyr")!')
        }
if(!require(lubridate, quietly = TRUE, warn.conflicts = FALSE)){
        stop('I need lubridate library. Please run install.packages("lubridate")!')
        }
if(!require(lattice, quietly = TRUE, warn.conflicts = FALSE)){
        stop('I need lattice library. Please run install.packages("lattice")!')
        }
   
if(!file.exists("activity.zip")){
        stop(paste('I need activity.zip file in the working directory.',
                  'Please copy activity.zip file into your working directory!',
                   sep = "\n"))
        }
```

Then, we download the dataset from the zip-file, convert it to `tbl_df`, and convert the data to a standard data format. Next we delete `df.activity` data frame since that frame is no loger needed. Finally, we group the loaded data and summarize them. The summary is represented with histogramin section 2. No `NA`s has been exluded from the analysis, yet. 

```{r, echo = TRUE}
df.activity     <- read.csv(unz("activity.zip", "activity.csv"), header = TRUE, stringsAsFactors = FALSE)
tbl.activity    <- tbl_df(df.activity)
tbl.activity    <- mutate(tbl.activity, date = ymd(date))

rm("df.activity")

tbl.by_date     <- group_by(tbl.activity,date)
tbl.total           <- summarize(tbl.by_date, totalperday = sum(steps, na.rm = TRUE))
```


## 2. Histogram of the total number of steps taken each day

```{r histogram1, fig.height = 4, echo = TRUE}
with(tbl.total, hist(totalperday,breaks = 8, xlab = "Total steps per day", main = "Histogram of total steps per day"))
```


## 3. Mean and median number of steps taken each day

For steps taken each day, the mean is `r mean(tbl.total$totalperday, na.rm = TRUE)`
and median is `r median(tbl.total$totalperday, na.rm = TRUE)`. 


## 4. Time series plot of the average number of steps taken (what is the average daily activity pattern?)

First we group and summarize the initial data by interval (we calculate the average number of steps per each interval).

```{r, echo = TRUE}
tbl.by_interval         <- group_by(tbl.activity, interval)
tbl.interval_average    <- summarize(tbl.by_interval, averageperinterval = mean(steps, na.rm = TRUE))
```

Then, we represent the result with the next plot. 

```{r plot4, echo = TRUE, fig.height = 4}
with(tbl.interval_average, 
     plot(interval, averageperinterval, type = "l", col = "blue", xlab = "5 minute interval no.", 
          ylab = "Average number of steps per interval"))

title(main = "Plot of average number of steps per interval")
```

## 5. The 5-minute interval that, on average, contains the maximum number of steps

The index of the 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps is `r arrange(tbl.interval_average, desc(averageperinterval))$interval[1]`, at which the average number of steps reaches its maximum that equals `r arrange(tbl.interval_average, desc(averageperinterval))$averageperinterval[1]`. 


## 6. Code to describe and show a strategy for imputing missing data

The total number of observation with `NA` in the dataset is `r sum(!complete.cases(tbl.activity))`.  

To fill in of the missing values, we substitute each of them with the mean value of average values for the corresponding interval. We take the average values from `tbl.interval_average`, which is already ready to use. 

```{r, echo = TRUE}
tbl.steps_imputed               <- tbl.activity
tbl.interval_average$subst_steps    <- sapply(tbl.interval_average$averageperinterval,  
                                          function(x) as.integer(round(x,0)))

tbl.steps_imputed               <- tbl.steps_imputed %>%
                                   merge(tbl.interval_average, by.x = "interval", by.y = "interval", all.x = TRUE) %>%
                                   mutate(steps = ifelse(is.na(steps), subst_steps, steps)) %>%
                                   select(steps,  date, interval) %>%
                                   arrange(date,interval)

tbl.imputed_by_date             <- group_by(tbl.steps_imputed,date)
tbl.total_imputed               <- summarize(tbl.imputed_by_date, totalperday = sum(steps, na.rm = TRUE))
```




## 7. Histogram of the total number of steps taken each day after missing values are imputed

The result of above transforations, grouping, and summazizing is represented with the next histogram. 

```{r histogram6, echo = TRUE, fig.height = 4}
with(tbl.total_imputed, hist(totalperday,breaks = 8, xlab = "Total steps per day", 
     main = "Histogram of total steps per day (na steps imputed)"))
```
    

For steps taken each day (imputed), the mean is `r mean(tbl.total_imputed$totalperday, na.rm = TRUE)`
and median is `r median(tbl.total_imputed$totalperday, na.rm = TRUE)`. Let's compare these results with the earlier ones, when `NA` were not imputed (numbers were rounded), inthe table below. 

```{r, echo=TRUE}
mean.init       <- trunc(mean(tbl.total$totalperday, na.rm = TRUE))
median.init     <- median(tbl.total$totalperday, na.rm = TRUE)

mean.imputed    <- trunc(mean(tbl.total_imputed$totalperday))
median.imputed  <- median(tbl.total_imputed$totalperday)

data.frame(row.names = c("NAs Remain","NAs Imputed", "Imp'd-Rem'n"), 
           Mean = c(mean.init, mean.imputed, mean.imputed - mean.init), Median = c(median.init,median.imputed, median.imputed - median.init))
```

The mean and median in the dataset, where `NA`s were imputed, are greater than those in the initial dataset. This is so because `NA`s were partially
substituded with strictly positive values that has made mean and median grow when switching consideration from the initial dataset to the dataset with
imputed values. 
    



Are there differences in activity patterns between weekdays and weekends?

## 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends (Are there differences in activity patterns between weekdays and weekends?)

```{r, echo = TRUE}
tbl.imputed_by_weekday.type_interval <- tbl.steps_imputed %>% 
                                        mutate(weekday.type = ifelse (between(wday(date), 2,6), 
                                              "weekday", "weekend")) %>%
                                        transform(weekday.type = factor(weekday.type)) %>%
                                        group_by(weekday.type, date, interval) %>%
                                        summarize(averagepertypedateinterval= mean(steps, na.rm = TRUE)) %>%
                                        group_by(weekday.type, interval) %>%
                                        summarize(averagepertypeinterval = mean(averagepertypedateinterval, 
                                                  na.rm = TRUE)) 
```


The result is represented with the following plot lattice:

```{r plot8, echo = TRUE, fig.height = 4}
xyplot(averagepertypeinterval ~ interval | weekday.type, data = tbl.imputed_by_weekday.type_interval, 
       layout = c(1, 2), type = 'l', xlab = "Interval", ylab = "Number of steps")
        
```

We can see some differences in activity patterns between weekdays and weekends:

- on weekdays, people beging walking earlier than on weekends
- on weekends, people walks more after the morning maximum as well as stop walking later than on weekdays
- on weekends, morning walking maximun is lower than on weekdays

We may assumme that people sleep longer and need not go so far on weekends in comparison to weekdays. Still, people have more *local* walk on weekends
because they have somewhat more spare time. 
