---
title: "Reproducible Research: Peer Assessment 1"
author: LexoL
date: January 10, 2016
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

## 1. Code for reading in the dataset and/or processing the data

```{r}
if(!require(dplyr, quietly = TRUE, warn.conflicts = FALSE)){
        stop('I need dplyr library. Please run install.packages("dplyr")!')
        }
if(!require(lubridate, quietly = TRUE, warn.conflicts = FALSE)){
        stop('I need lubridate library. Please run install.packages("lubridate")!')
        }
if(!require(lattice, quietly = TRUE, warn.conflicts = FALSE)){
        stop('I need lattice library. Please run install.packages("lattice")!')
        }
   
if(!file.exists("activity.zip")){
        stop(paste('I need activity.zip file in the working directory.',
                  'Please copy activity.zip file into your working directory!',
                   sep = "\n"))
        }


df.activity     <- read.csv(unz("activity.zip", "activity.csv"), header = TRUE, stringsAsFactors = FALSE)
tbl.activity    <- tbl_df(df.activity)
tbl.activity    <- mutate(tbl.activity, date = ymd(date))

rm("df.activity")

tbl.by_date     <- group_by(tbl.activity,date)
total           <- summarize(tbl.by_date, totalperday = sum(steps, na.rm = TRUE))
```

## Histogam 2
2. Histogram of the total number of steps taken each day

```{r histogram1, fig.height = 4}
with(total, hist(totalperday,breaks = 8, xlab = "Total steps per day", main = "Histogram of total steps per day"))
```

## What is mean total number of steps taken per day?
3. Mean and median number of steps taken each day

For steps taken each day, the mean is `r mean(total$totalperday, na.rm = TRUE)`
and median is `r median(total$totalperday, na.rm = TRUE)`. 


## What is the average daily activity pattern?
4. Time series plot of the average number of steps taken

```{r}
tbl.by_interval         <- group_by(tbl.activity, interval)
interval_average        <- summarize(tbl.by_interval, averageperinterval = mean(steps, na.rm = TRUE))
```

Draw plot

```{r plot4, fig.height = 4}
with(interval_average, 
     plot(interval, averageperinterval, type = "l", col = "blue", xlab = "5 minute interval no.", 
          ylab = "Average number of steps per interval"))

title(main = "Plot of average number of steps per interval")
```

5. The 5-minute interval that, on average, contains the maximum number of steps

The index of the 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps is `r arrange(interval_average, desc(averageperinterval))$interval[1]`, at which the the average number of steps reach maximum that equals `r arrange(interval_average, desc(averageperinterval))$averageperinterval[1]`. 


## Imputing missing values
6. Code to describe and show a strategy for imputing missing data

The total number of observation with NA in the dataset is `r sum(!complete.cases(tbl.activity))`.

To fill in of the missing values, we substitute each of them with the mean value of average values for the corresponding interval. 

```{r}
tbl.steps_imputed               <- tbl.activity
interval_average$subst_steps    <- sapply(interval_average$averageperinterval,  
                                          function(x) as.integer(round(x,0)))

tbl.steps_imputed               <- tbl.steps_imputed %>%
                                   merge(interval_average, by.x = "interval", by.y = "interval", all.x = TRUE) %>%
                                   mutate(steps = ifelse(is.na(steps), subst_steps, steps)) %>%
                                   select(steps,  date, interval) %>%
                                   arrange(date,interval)

tbl.imputed_by_date             <- group_by(tbl.steps_imputed,date)
total_imputed                   <- summarize(tbl.imputed_by_date, totalperday = sum(steps, na.rm = TRUE))
```

7. Histogram of the total number of steps taken each day after missing values are imputed

Histogram of the total number of steps taken each day (imputed)

```{r histogram6, fig.height = 4}
with(total_imputed, hist(totalperday,breaks = 8, xlab = "Total steps per day", 
     main = "Histogram of total steps per day (na steps imputed)"))
```
    
HEADING!!! INTERPRETATION OF CHANGES    

Mean and median number of steps taken each day (imputed)   

For steps taken each day (imputed), the mean is `r mean(total_imputed$totalperday, na.rm = TRUE)`
and median is `r median(total_imputed$totalperday, na.rm = TRUE)`. 



## Are there differences in activity patterns between weekdays and weekends?

8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}
tbl.imputed_by_weekday.type_interval <- tbl.steps_imputed %>% 
                                        mutate(weekday.type = ifelse (between(wday(date), 2,6), 
                                              "weekday", "weekend")) %>%
                                        transform(weekday.type = factor(weekday.type)) %>%
                                        group_by(weekday.type, date, interval) %>%
                                        summarize(averagepertypedateinterval= mean(steps, na.rm = TRUE)) %>%
                                        group_by(weekday.type, interval) %>%
                                        summarize(averagepertypeinterval = mean(averagepertypedateinterval, 
                                                  na.rm = TRUE)) 
```

Draw the plot

```{r plot8, fig.height = 4}
xyplot(averagepertypeinterval ~ interval | weekday.type, data = tbl.imputed_by_weekday.type_interval, 
       layout = c(1, 2), type = 'l', xlab = "Interval", ylab = "Number of steps")
        
```
