---
title: "Reproducible Research: Peer Assessment 1"
author: LexoL
date: January 10, 2016
output: 
  html_document:
    keep_md: true
---

# TO DISCARD
tbl.activity <- mutate(tbl.activity, date = ymd(date), wd = as.character(wday(date, label = TRUE)), weekday.type = ifelse (between(wday(date), 2,6), "weekday", "weekend"))

1. Code for reading in the dataset and/or processing the data
2. Histogram of the total number of steps taken each day
3. Mean and median number of steps taken each day
4. Time series plot of the average number of steps taken
5. The 5-minute interval that, on average, contains the maximum number of steps
6. Code to describe and show a strategy for imputing missing data
7. Histogram of the total number of steps taken each day after missing values are imputed
8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
9. All of the R code needed to reproduce the results (numbers, plots, etc.) in the report
 
## Loading and preprocessing the data

1. Code for reading in the dataset and/or processing the data

```{r}
library(dplyr)
library(lubridate)

df.activity <- read.csv(unz("activity.zip", "activity.csv"), header = TRUE, stringsAsFactors = FALSE)
tbl.activity <- tbl_df(df.activity)
rm("df.activity")
tbl.activity <- tbl.activity %>%
        mutate(date = ymd(date), weekday.type.char = ifelse (between(wday(date), 2,6), "weekday", "weekend")) %>%
        transform(weekday.type = factor(weekday.type.char))
tbl.by_date <- group_by(tbl.activity,date)
total <- summarize(tbl.by_date, totalperday = sum(steps, na.rm = TRUE))
```

# Histogam 2
2. Histogram of the total number of steps taken each day

```{r histogram1, fig.height = 4}
with(total, hist(totalperday,breaks = 8, xlab = "Total steps per day"))

```

## What is mean total number of steps taken per day?
3. Mean and median number of steps taken each day

For steps taken each day, the mean is `r mean(total$totalperday, na.rm = TRUE)`
and median is `r median(total$totalperday, na.rm = TRUE)`. 




## What is the average daily activity pattern?
4. Time series plot of the average number of steps taken

```{r}
tbl.by_interval <- group_by(tbl.activity, interval)
interval_average <- summarize(tbl.by_interval, averageperinterval = mean(steps, na.rm = TRUE))
```

Draw plot

```{r plot4, fig.height = 4}
with(interval_average, plot(interval, averageperinterval, type = "l", 
                            col = "blue", xlab = "5 minute interval no.", 
                            ylab = "Average number of steps per interval"))

```




## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
